HO unification from object language to meta language

### 2024-03-15 paper/main.tex
latex

\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress XXX improve....

We represent a logic program run for a goal g as
a list steps $p$ of length $\mathcal{N}$ each made of a unification problem
between terms $\mathcal{S}_{p_l}$ and $\mathcal{S}_{p_r}$
taken from the set of all terms $\mathcal{S}$.
If a rule is used
twice in a run we make two copies of the terms (the predicates)
occurring in the head or premises. So a run is the conjunction of all
unification problems, where $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepFD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runHD}[3]{\ensuremath{\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{\sigma'}{\sigma_{p}}{#3}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho'}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate this run as where $\sigma_0$ is the empty substitution
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
  \qquad \runFD{\mathcal{S}}{\mathcal{N}}{\mathbb{L}}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\rho_{\mathcal{N}}}
\end{array}
$$

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of \hrun, we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}

\newcommand{\maybebeta}{\ensuremath{\stackrel{?}{\beta}}\xspace}
\begin{definition}[\maybebeta]
  $t$ is in \maybebeta if it contains a subterm matching the
  pattern \elpiIn{app[uva _ _|_]}
\end{definition}

\newcommand{\maybeeta}{\ensuremath{\stackrel{?}{\eta}}\xspace}
\begin{definition}[\maybeeta]
  $t$ is in \maybeeta if it contains a subterm matching the
  pattern \elpiIn{lam x\ .. lam y\ app[con _,x, ...,uva _ [y]]}\\
   or
  \elpiIn{lam x\lam y\uva _ [y,x]}
\end{definition}

@@

\begin{proposition}[Normal form preservation]
  $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
  \Rightarrow norm(\sigma\mathcal{T})$   
\end{proposition}

If we focus on a single unification problem, that is
$\mathcal{S} = \{ s_1, s_2 \}$ and 
$\C{s_1}{t_1}{m_1}{l_1}$ and $\C{s_2}{t_2}{m_2}{l_2}$ we write
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land
    \D{\sigma''}{\{l_1,l_2\}}{\rho}
\end{array}
$$
\noindent
We write $s \in \llambda$ if all unif variables in
$s$ are applied to distinct bound variables.

@@

\paragraph{The intuition} is that a term s is compiled in a term t where any
``problematic'' sp is replaced by a hole hp and an accessory
(suspended) unification problem tp \Ue hp (called link) where
tp is the compiled of sp. 
On t \Ue is well behaved (captures \Eo), and check solves links when it can by
using \Ue or keeps them suspended until the decompilation phase.
check also detects occur checks trough the suspended unif problems
in order to grant \cref{prop:fidelity}.

>>>>

\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress XXX improve....

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \solve(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho'}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \solve on the set of links. We claim the following:

\begin{proposition}[Simulation]\label{prop:sumul}
$\forall \mathcal{S}, \forall \mathcal{N}$
$$
  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
$$
\end{proposition}

\noindent
That is, the two executions give the same result. Moreover:

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}
\noindent
Note that this property guarantees that a failure in the \Fo run
is matched by a failure
in \Ho
\emph{at the same step}.

\newcommand{\maybebeta}{\ensuremath{\beta}\xspace}
\begin{definition}[\maybebeta]
  $t$ is in \maybebeta if it contains a subterm matching the
  pattern \elpiIn{app[uva _ _|_]}
\end{definition}

\newcommand{\maybeeta}{\ensuremath{\eta}\xspace}
\begin{definition}[\maybeeta]
  $t$ is in \maybeeta if it contains a subterm matching the
  pattern \elpiIn{lam x\ .. lam y\ app[con _,x, ...,uva _ [y]]}\\
   or
  \elpiIn{lam x\lam y\uva _ [y,x]}
\end{definition}

@@

\begin{proposition}[Normal form preservation]\label{prop:nf}
  $norm(\sigma\mathcal{T}) \land \mathrm{step'}(p,\sigma) \mapsto \sigma'
  \Rightarrow norm(\sigma\mathcal{T})$   
\end{proposition}

TODO: check if this is also true for $\rho$ after decomp.
In particular this guaarantees that is we start from normal terms
we never introduce eta-long or non-beta-normal terms. This is required in
order to show the fidelity, since there terms would not be considered
as equal by \Ue.

If we focus on a single unification problem, that is
$\mathcal{S} = \{ s_1, s_2 \}$ we can define \Uo as follows:
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$
\noindent
Note that proposition \ref{prop:nf} does not hold in this case since
decompilation can introduce terms in \maybebeta or \maybebeta. Nevertheless
\Uo has the following properties (where we
write $s \in \llambda$ if all unif variables in
$s$ are applied to distinct bound variables):

@@

\subsection{The intuition in a nutshell}
A term $s$ is compiled in a term $t$ where any
``problematic'' subterm $tp$ is replaced by a unification variable
$hp$ and an accessory
(suspended) unification problem $(hp \Ue tp) \in \mathcal{L}$.
Since $t \in \llambda$, \Ue is well behaved (captures \Eo).

Problematic terms are terms where the head constructor may
change if instantiated, that is, they are in \maybebeta or \maybeeta.
On these terms \Ue is weak/incorrect.
\begin{itemize}
\item \elpiIn{app[F,con "a"]} may be \Eo to \elpiIn{con "a"}
\item \elpiIn{lam x\lam y\app[F,y,x]} may be \Eo to \elpiIn{con "f"} 
\end{itemize}

\noindent
\solve{} solves links when they are no more ``problematic'', for example
in the examples above if \elpiIn{F} materializes, or if the
other side of the link forces a commit. EXPLAIN.
In these cases the problems are ``resumed'' by using \Ue,
otherwise they are kept suspended until the
decompilation phase.
\solve also detects occur checks trough the suspended unif problems
in order to grant \cref{prop:fidelity}.

Note tha the first case is outside $\llambda$ but is frequent in practice
so we handle it. It is also what prop 5 is about.
